---
title: "BIOL-A601 R Assignment 3"
author: "Timm Nawrocki"
date: "November 8, 2018"
output: 
  html_document: 
    df_print: kable
    number_sections: yes
    self_contained: no
    toc: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

# Code header

```{r header}
# -*- coding: utf-8 -*-
# ---------------------------------------------------------------------------
# R Assignment 3: Data Summary
# Author: Timm Nawrocki
# Created on: 2018-10-13
# Usage: Code is intended to be executed as an R markdown in RStudio.
# Description: "BIOL-A601 R Assignment 3" includes examples for normal and non-normal distributions.
# ---------------------------------------------------------------------------
```

# Set up session

The set up code installs libraries used in this assignment if any are not already installed. Sets a working file based on user input.

```{r environment}
# Install required libraries if they are not already installed
Required_Packages <- c("dplyr", "ggplot2", "ggpmisc", "readxl", "tidyr", "pander", "gridExtra")
New_Packages <- Required_Packages[!(Required_Packages %in% installed.packages()[,"Package"])]
if (length(New_Packages) > 0) {
  install.packages(New_Packages)
}

# Import required libraries
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(readxl)
library(tidyr)
library(pander)
library(gridExtra)

# Select input files
protein_file = file.choose()
```

## Define functions

```{r functions}
# Define a function to generate standardized histograms
histogram = function(inData, variable, interval, range, title, x_label, y_label) {
  font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
  normalDist = function(x, mean, sd, n, interval) {
    dnorm(x = x, mean = mean, sd = sd) * n * interval
    }
  plot = ggplot(data=inData, aes(variable)) +
    theme_bw() +
    geom_histogram(aes(y=..count..),
                   breaks=seq(range[1], range[2], by = interval),
                   colour = 'white',
                   fill = 'grey20',
                   alpha = 0.5) +
    stat_function(fun = normalDist,
                  args = c(mean = mean(variable),
                           sd = sd(variable),
                           n = length(variable),
                           interval = interval)) +
    labs(title=title) +
    labs(x=x_label, y=y_label) +
    font_size
  return(plot)
}

# Define a function to generate standardized box plots
box_plot = function(inData, X, y, title, x_label, y_label) {
  font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
  ggplot(data=inData, aes(X, y)) +
    geom_boxplot() +
    theme_bw() +
    labs(title=title) +
    labs(x=x_label, y=y_label) +
    font_size
}

# Define a function to generate qq plots
qq_plot = function(inData, variable, title, x_label, y_label) {
  font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
  y = quantile(variable, c(0.25, 0.75), type=5)
  x = qnorm(c(0.25, 0.75))
  slope = diff(y) / diff(x)
  int = y[1] - slope * x[1]
  ggplot() + aes(sample=variable) +
    stat_qq(distribution=qnorm) +
    geom_abline(intercept=int, slope=slope) +
    theme_bw() +
    labs(title=title) +
    labs(x=x_label, y=y_label) +
    font_size
}


# Define a function to report 95% confidence interval lower and upper bounds
confidence = function(mean, sd, n) {
  error = qt(0.975, df=sample_n-1) * sample_sd / sqrt(sample_n)
  upper = mean + error
  lower = mean - error
  return(c(upper, lower))
}
```

# Explore Salix pulchra protein data

```{r protein}
# Create data frame from Salix pulchra protein csv file
protein = read.csv(protein_file, header=TRUE, stringsAsFactors=FALSE)
protein$Site = recode(protein$Site, DNP = "Denali", NSA = "Nelchina")

# Summarize protein data by site
mean_site = protein %>%
  group_by(Site) %>%
  summarize(num_obs= n(),
            mean_n = signif(mean(Nconc), digits=3),
            sd_n = signif(sd(Nconc), digits=3),
            round(sd(Nconc)/mean(Nconc)*100, digits=1))

# Create human readable column names for table display
colnames(mean_site)[1] = 'Site'
colnames(mean_site)[2] = 'Number'
colnames(mean_site)[3] = 'Mean nitrogen conc.'
colnames(mean_site)[4] = 'Std dev nitrogen conc.'
colnames(mean_site)[5] = 'Coeficient of variation (%)'

# Create output table
mean_site
```

## Visualize data as box plot

```{r boxPlot}
# Create a box plot for nitrogen concentration by site
protein_boxplot = box_plot(protein,
                           protein$Site,
                           protein$Nconc,
                           "Box plot of nitrogen concentration by site",
                           "Site",
                           "Nitrogen concentration")
protein_boxplot
```

## Visualize Denali and Nelchine data as histograms

```{r histograms, fig.height = 5, fig.width = 14, fig.align = "center"}
# Subset the protein data to include only observations for Denali
denali_protein = subset(protein, Site=='Denali')

# Create a histogram plot for nitrogen concentration for Denali
denali_histogram = histogram(denali_protein,
                             denali_protein$Nconc,
                             0.4,
                             c(1.4, 3.4),
                             "Histogram of nitrogen concentrations in Denali National Park",
                             "Nitrogen concentration",
                             "Count")


# Subset the protein data to include only observations for Nelchina
nelchina_protein = subset(protein, Site=='Nelchina')

# Create a histogram plot for nitrogen concentration for Nelchina
nelchina_histogram = histogram(nelchina_protein,
                             nelchina_protein$Nconc,
                             0.4,
                             c(0.8, 3.6),
                             "Histogram of nitrogen concentrations in Denali National Park",
                             "Nitrogen concentration",
                             "Count")

plot_grid = grid.arrange(denali_histogram, nelchina_histogram, nrow=1)
```

## Test normality of Denali data

```{r denaliNormal}
# Run a Shapiro-Wilk test on the Denali nitrogen concentrations
denali_shapiro = shapiro.test(denali_protein$Nconc)

# Create a qq plot for the denali nitrogen concentrations
protein_qq = qq_plot(denali_protein,
                     denali_protein$Nconc,
                     "Normal Q-Q plot test for Denali nitrogen concentration",
                     "Theoretical distribution",
                     "Nitrogen concentration")
protein_qq
cat('The distribution of the Denali data is statistically indistinguishable from a normal distribution:')
denali_shapiro
```

# Test the effects of sample size on accuracy of Shapiro-Wilk test

Output a plot showing the relationship between log2(n) and percent of correct determinations by Shapiro-Wilk test with loess smoothed conditional mean and 95% confidence interval.

```{r accuracy}
# Set a sample size counter
n = 5
# Create empty lists to store results
n_list = c()
percent_correct = c()
# Define a function to test a sample size 100 times and return a percent correct
shapiro_accuracy = function(n, percent_correct){
  i = 1
  correct_results = c()
  while (i < 101) {
    distribution = rbeta(n, 10, 100)
    distribution_shapiro = shapiro.test(distribution)
    if (distribution_shapiro[2] <= 0.05) {
      result = 1
    } else {
      result = 0
    }
    correct_results = c(correct_results, result)
    i = i+1
  }
  accuracy = mean(correct_results)
  percent_correct = c(percent_correct, accuracy)
  return(percent_correct)
}
# Iterate through 10 increasing sample sizes and calculate percent correct for each
while (n < 2600) {
  n_list = c(n_list, n)
  percent_correct = shapiro_accuracy(n, percent_correct)
  n = n*2
}
# Store results to data frame
n_list = log2(n_list)
percent_correct = percent_correct*100
shapiro_results = cbind.data.frame(n_list, percent_correct)
# Plot results
font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
shapiro_plot = ggplot(data=shapiro_results, aes(x=n_list, y=percent_correct)) +
  geom_point() +
  geom_smooth(method='loess',
                color="black",
                fill="grey20",
                size=0.5) +
  theme_bw() +
  labs(title="Percent correct determinations by Shapiro Test") +
  labs(x="Sample size (log2)", y="Percent correct (%)") +
  font_size
shapiro_plot
cat('Percent correct determined by 100 independent distributions per sample size.')
```