---
title: 'R Assignment 2: Data Summary'
author: "Timm Nawrocki"
date: "September 24, 2018"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

## Code header

```{r header}
# -*- coding: utf-8 -*-
# ---------------------------------------------------------------------------
# R Assignment 2: Data Summary
# Author: Timm Nawrocki
# Created on: 2018-09-09
# Usage: Code is intended to be executed as an R markdown or R script in RStudio.
# Description: "R Assignment 2: Data Summary" is an R assignment for BIOL-A601, Fall 2018.
# ---------------------------------------------------------------------------
```

## Set up session environmental settings

**Note:** Install libraries used in this assignment if not already installed. Set user working directory based on user input.

```{r environment}
# Install required libraries if they are not already installed
Required_Packages <- c("dplyr", "ggplot2", "ggpmisc", "readxl", "tidyr", "pander")
New_Packages <- Required_Packages[!(Required_Packages %in% installed.packages()[,"Package"])]
if (length(New_Packages) > 0) {
  install.packages(New_Packages)
}

# Import required libraries
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(readxl)
library(tidyr)
library(pander)

# Select input files
seeds_file = file.choose()
daphnia_file = file.choose()
```
## Define functions

```{r functions}
# Define a function to generate standardized histograms
histogram = function(inData, variable, interval, range, title, x_label, y_label) {
  font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
  normalDist = function(x, mean, sd, n, interval) {
    dnorm(x = x, mean = mean, sd = sd) * n * interval
    }
  plot = ggplot(data=inData, aes(variable)) +
    theme_bw() +
    geom_histogram(aes(y=..count..),
                   breaks=seq(range[1], range[2], by = interval),
                   colour = 'white',
                   fill = 'cornflowerblue',
                   alpha = 0.5) +
    stat_function(fun = normalDist,
                  args = c(mean = mean(variable),
                           sd = sd(variable),
                           n = length(variable),
                           interval = interval)) +
    labs(title=title) +
    labs(x=x_label, y=y_label) +
    font_size
  return(plot)
}

# Define a function to generate standardized box plots
box_plot = function(inData, X, y, title, x_label, y_label) {
  font_size = theme(axis.title = element_text(size=14), axis.text = element_text(size = 12))
  ggplot(data=inData, aes(X, y)) +
    geom_boxplot() +
    theme_bw() +
    labs(title=title) +
    labs(x=x_label, y=y_label) +
    font_size
}

# Define a function to report 95% confidence interval lower and upper bounds
confidence = function(mean, sd, n) {
  error = qt(0.975, df=sample_n-1) * sample_sd / sqrt(sample_n)
  upper = mean + error
  lower = mean - error
  return(c(upper, lower))
}
```

## Explore hemlock data

```{r dataImport}
# Create data frame from seeds csv file
seeds = read.csv(seeds_file, header=TRUE, stringsAsFactors = FALSE)
```

### Summarize data by site

```{r simpleSummary}
# Summarize the seeds dataset by site for number of observations, mean weight, and coefficient of variation for weight
mean_site = seeds %>%
  group_by(site) %>%
  summarize(num_obs = n(), mean_weight = signif(mean(seed_wt), digits=2), site_cv = round(sd(seed_wt)/mean(seed_wt)*100, digits=0))

# Create human readable column names for table display
colnames(mean_site)[1] = 'Site'
colnames(mean_site)[2] = 'Number'
colnames(mean_site)[3] = 'Mean Weight (mg)'
colnames(mean_site)[4] = 'Coefficient of Variation (%)'

# Create output table
pander(mean_site)
```

### Histogram

```{r histogram}
# Create a histogram plot for seed frequency by intervals of 0.0005 mg for all sites
seed_histogram = histogram(seeds, seeds$seed_wt, 0.0005, c(0, 0.005), "Histogram of Seed Weight", "Seed Weight (mg)", "Count")
seed_histogram
```

### Box plot by site

```{r boxplot}
# Create a box plot for seed weight by site
seed_boxplot = box_plot(seeds, seeds$site, seeds$seed_wt, "Box plot of seed weight by site", "Site", "Seed Weight (mg)")
seed_boxplot
```

### Confidence interval

```{r confidence_interval}
# Compute sample mean, standard deviation, and size
sample_mean = mean(seeds$seed_wt)
sample_sd = sd(seeds$seed_wt)
sample_n = length(seeds$seed_wt)

# Calculate the 95% confidence interval
confidence_interval = confidence(sample_mean, sample_sd, sample_n)

# Report the coefficient of variation for seed weight and upper and lower bounds of 95% confidence interval
cat(paste('Coefficient of variation = ', toString(round(sample_sd/sample_mean*100, digits=0)), '%\n', 'Upper bound of 95% confidence interval is ', toString(signif(confidence_interval[1], digits=2)), ' mg\n', 'Lower bound of 95% confidence interval is ', toString(signif(confidence_interval[2], digits=2)), ' mg\n', sep=''))
```

## Explore hypothetical fish population

```{r fish_stats}
# Define a hypothetical population of 1,000 fish with mean length of 50 cm and standard deviation in length of 5 cm in a normal distribution
length_mean = 50
length_sd = 5
fish_n = 1000
fish_population = rnorm(fish_n, length_mean, length_sd)
fish_population = data.frame(fish_population)

# Create a histogram of fish length
fish_histogram = histogram(fish_population, fish_population$fish_population, 4, c(30, 70), "Histogram of Fish Length", "Length (cm)", "Count")
fish_histogram

# Calculate mean and standard deviation
cat(paste('Mean = ', toString(round(mean(fish_population$fish_population), digits=1)), ' cm\n', 'Standard deviation = ', toString(round(sd(fish_population$fish_population), digits=1)), ' cm\n', sep=''))
```

## Sample from normally distributed data

**Note:** I am not sure what the suggested experiment from Lecture 3 Slide 6 is because I don't see the lecture on blackboard. Instead I've just plotted a sample from a hypothetical population of normally distributed data.

```{r sample}
# Create a hypothetical population of count data
mean = 232
sd = 34
n = 10000
population = rnorm(n, mean, sd)
sample_population = sample(population, 100)
population = data.frame(population)
sample_population = data.frame(sample_population)

# Create a plot of full population against the normal distribution curve
full_plot = histogram(population, population$population, 20, c(100, 400), "Histogram of Entire Normally Distributed Population", "Hypothetical X (unit)", "Count")
full_plot

# Create a plot of subsample against the normal distribution curve
sample_plot = histogram(sample_population, sample_population$sample_population, 20, c(100, 400), "Histogram of 1% Sample of Population", "Hypothetical X (unit)", "Count")
sample_plot

```

## Explore "Daphnia" Dataset

```{r loadFile}
# Create data frame from Daphnia excel file
river_data = read_excel(daphnia_file)
```

### Summarize all data

```{r bothRivers}
# Calculate mean, standard deviation, coefficient of variation, and upper and lower bounds of 95% confidence interval
mean_gr = mean(river_data$Growth.rate)
sd_gr = sd(river_data$Growth.rate)
cv_gr = sd_gr/mean_gr*100
upper_gr = confidence(mean_gr, sd_gr, length(river_data$Growth.rate))[1]
lower_gr = confidence(mean_gr, sd_gr, length(river_data$Growth.rate))[2]

# Report results
cat(paste('Mean = ', toString(round(mean_gr, digits=2)), '\n', 'Standard deviation = ', toString(round(sd_gr, digits=2)), '\n', 'Coefficient of Variation = ', toString(round(cv_gr, digits=0)), '%\n', 'Upper bound of 95% confidence interval = ', toString(round(upper_gr, digits=4)), '\n', 'Lower bound of 95% confidence interval = ', toString(round(lower_gr, digits=4)), '\n', sep=''))
```

### Summarize data by river

```{r perRiver}
# Summarize the growth rate dataset by river for number of observations, mean weight, coefficient of variation, upper 95% confidence interval, and lower 95% confidence interval
mean_river = river_data %>%
  group_by(River) %>%
  summarize(num_obs = n(), mean_rate = signif(mean(Growth.rate), digits=4), river_cv = round(sd(Growth.rate)/mean(Growth.rate)*100, digits=1), upper_ci = round(confidence(mean(Growth.rate), sd(Growth.rate), length(Growth.rate))[1], digits=5), lower_ci = round(confidence(mean(Growth.rate), sd(Growth.rate), length(Growth.rate))[2], digits=5))

# Create human readable column names for table display
colnames(mean_river)[1] = 'River'
colnames(mean_river)[2] = 'Number'
colnames(mean_river)[3] = 'Mean growth rate'
colnames(mean_river)[4] = 'Coefficient of Variation (%)'
colnames(mean_river)[5] = 'Upper CI'
colnames(mean_river)[6] = 'Lower CI'

# Create output table
pander(mean_river)
```

### Histogram for River Tyne

```{r tyneHistogram}
# Subset the river data to include only observations for the Tyne
tyne_data = subset(river_data, River=='Tyne')

# Create a histogram plot for growth rate frequency for the Tyne
tyne_histogram = histogram(tyne_data, tyne_data$Growth.rate, 0.5, c(2, 6), "Histogram of Growth Rates on the Tyne River", "Growth rate", "Count")
tyne_histogram
```

### Histogram for River Wear

```{r wearHistogram}
# Subset the river data to include only observations for the Wear
wear_data = subset(river_data, River=='Wear')

# Create a histogram plot for growth rate frequency for the Wear
wear_histogram = histogram(wear_data, wear_data$Growth.rate, 0.5, c(1, 7.5), "Histogram of Growth Rates on the Wear River", "Growth rate", "Count")
wear_histogram
```

### Box plot by river

```{r riverBoxplot}
# Create a box plot for growth rate by river
river_boxplot = box_plot(river_data, river_data$River, river_data$Growth.rate, "Box plot of growth rates by river", "River", "Growth rate")
river_boxplot
```