---
title: "Reproducible Analyses Workshop"
author: "Timm Nawrocki"
date: "May 17, 2018"
output:
  html_document:
    keep_md: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This repository is a training repository with notes from the May 2018 SASAP Reproducible Research training.

## Getting Started
These instructions will enable you to run the training resource scripts. This readme file is the output of the "ReproducibleMarkdown.rmd" script.

## Prerequisites
1. R 3.5.0+
2. RStudio 1.1.453+

## Installing
1. The repository should be copied to a local drive. The script files are designed to be run in RStudio.

## Usage

```{r Set Libraries, include=FALSE}
# Install required libraries if they are not already installed.
Required_Packages <- c("devtools", "dplyr", "DT", "ggplot2", "ggpmisc", "kableExtra", "leaflet", "readxl", "remotes", "tidyr")
New_Packages <- Required_Packages[!(Required_Packages %in% installed.packages()[,"Package"])]
if (length(New_Packages) > 0) {
  install.packages(New_Packages)
}

# Import required libraries.
library(devtools)
library(dplyr)
library(DT)
library(ggplot2)
library(ggpmisc)
library(kableExtra)
library(leaflet)
library(readxl)
library(remotes)
library(tidyr)

# Set seed
set.seed(415)

# Set working directory by user input
userDirectory <- choose.dir()
setwd(userDirectory)
```

### Environment Settings

The environment settings for this session are:

```{r environmentSettings, echo=FALSE}
# Print session info
sessionInfo()
```

### Example Plot

The following code produces an example point plot based on random numbers.

```{r examplePlot}
ggplot(data.frame(x=rnorm(100), y=rnorm(100)), aes(x,y)) +
  geom_point()
data('mpg')
```

### Example Table

The following code produces an example table using the "mpg" dataset.

```{r datatable}
data('mpg')
datatable(mpg, caption='Mileage')
```

## Credits

### Built With
* R 3.5.0
* RStudio 1.1.453

### Authors
* **Timm Nawrocki** - *Alaska Center for Conservation Science, University of Alaska Anchorage*

### Usage Requirements
* none

### License
This project is private for my notes.