---
title: "Reproducible Analyses Workshop"
author: "Timm Nawrocki"
date: "May 17, 2018"
output:
  html_document:
    keep_md: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Set Libraries, include=FALSE}
# Install required libraries if they are not already installed.
Required_Packages <- c("devtools", "dplyr", "DT", "ggplot2", "ggpmisc", "kableExtra", "leaflet", "readxl", "remotes", "tidyr")
New_Packages <- Required_Packages[!(Required_Packages %in% installed.packages()[,"Package"])]
if (length(New_Packages) > 0) {
  install.packages(New_Packages)
}

# Import required libraries.
library(devtools)
library(dplyr)
library(DT)
library(ggplot2)
library(ggpmisc)
library(kableExtra)
library(leaflet)
library(readxl)
library(remotes)
library(tidyr)

# Set seed
set.seed(415)

# Set working directory by user input
userDirectory <- choose.dir()
setwd(userDirectory)
```

## Environment Settings

The environment settings for this session are:

```{r environmentSettings, echo=FALSE}
# Print session info
sessionInfo()
```

## Example Plot

The following code produces an example point plot based on random numbers.

```{r examplePlot}
ggplot(data.frame(x=rnorm(100), y=rnorm(100)), aes(x,y)) +
  geom_point()
data('mpg')
```

## Example Table

The following code produces an example table using the "mpg" dataset.

```{r datatable}
data('mpg')
datatable(mpg, caption='Mileage')
```

## Tidying data using dplyr and tidyr

```{r readCSV}
# Read in csv from url
catch_data = read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"), stringsAsFactors = FALSE)
```

The pipe function takes an initial defined argument and passes it as the first argument in an initial command. For every subsequent command, the output from the last command is input as the first argument in a subsequent command. The final command needs to be stored as a variable.

```{r gatherData}
# Select columns for analysis and transpose into a new table shape.
catch_data = catch_data %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

# Transpose the data frame and save as a new data frame.
catch_gather = gather(catch_data, key = species, value = catch, Chinook, Sockeye, Coho, Pink, Chum)
```

The output table has "gathered" the catch-species columns into a "species" column and a "catch" column. The key specifies the new column name for the column names listed in the arguments and the value specifies the new column name for the values that existed within the columns listed in the arguments.

```{r spreadData}
catch_spread = spread(catch_gather, key = species, value = catch, fill = NA)
```

The syntax for spreading the data functions in opposite such that key specifies the column headings to be added and the value parses into those columns.

```{r catchError}
# Add a new column to data frame and multiply by 1000 to correct for units. An error for a non-numeric value in row 401 is corrected by converting to integer.
catch_gather <- catch_gather %>% 
  rename(catch_thousands = catch) %>%
  mutate(catch_thousands = ifelse(catch_thousands == "I", 1, catch_thousands),
         catch_thousands = as.integer(catch_thousands)) %>%
  mutate(catch = catch_thousands * 1000)
```